Concurrent Programming with CXL

### 2020-05-20 book/paper.tex
Starting semaphore chapter
Added correctness arguments for mutual exclusion in the otherwise broken solutions.

Figure~\ref{fig:upflags} presents a solution based on each process having
a flag indicating that it is trying to enter the critical section.
A process can write its own flag and read the flag of its peer.
After setting its flag, the process waits until the other process
(\texttt{1 - self}) is not trying to enter the critical section.
If we run this program, the assertion does not fail.  In fact, this
solution does prevent both processes being in the critical section at
the same time.
However, it has a different problem called \emph{livelock}: if both try to
enter the critical section at the same time, they may end up waiting for
one another.

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
The solution has the nice properties that it satifies mutual exclution and,
moreover, processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

\chapter{Local Variables}

>>>>

Figure~\ref{fig:upflags} presents a solution based on each process having
a flag indicating that it is trying to enter the critical section.
A process can write its own flag and read the flag of its peer.
After setting its flag, the process waits until the other process
(\texttt{1 - self}) is not trying to enter the critical section.
If we run this program, the assertion does not fail.  In fact, this
solution does prevent both processes being in the critical section at
the same time.

To see why, first note the invariant that if process $i$ is in the
critical section, then flags[i] == True.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not at the critical section at time $t_1$.
Suppose that process~1 is at the critical section from $t_2$ until
it clears its flag at time $t_3$.  Note $t_0 < t_1 < t_2 < t_3$.
Process~1 can only reach the critical section between $(t_2, t_3)$
if it finds \texttt{flags[0] == False} during $(t_1, t_3)$.  But
during $(t_0, t_3)$ \texttt{flags[0] == True}.

However, the solution has a different problem called \emph{livelock}:
if both try to enter the critical section at the same time, they may
end up waiting for one another indefinitely.  Thus the solution
violated \emph{progress}.

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process $i$ is in the critical
section, \texttt{turn} == $i$.  Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it also violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

\chapter{Semaphores}
